---
title: "wOBA and wRC+"
author: "Frida Gomam"
date: 2015-07-23T21:13:14-05:00
categories: ["R"]
tags: ["sabermetrics"]
weight: 1
---
```{r setup, include = FALSE}
knitr::opts_knit$set(root.dir = "/Users/walkerharrison/Desktop/Website/data/chadwick-0.6.5/")
knitr::opts_chunk$set(warning=FALSE,
                      message=FALSE,
                      cache = TRUE)

```


The first step, naturally, is to read in some data. If you've used retrosheet event files before, the `get_plays` function will check if the right .csv file already exists, and use the aforementioned `parse.retrosheet2.pbp` function if it doesn't. Once the column names are attached, we have our raw play-by-play data for a single season.

```{r}
library(tidyverse)
library(gridExtra)
source("parse.retrosheet.pbp.R")

get_plays <- function(season){

  season.path <- paste0('download.folder/unzipped/all', season, '.csv')
  if (!file.exists(season.path)) {parse.retrosheet2.pbp(season)}
  
  plays <- read_csv(season.path)
  fields <- read_csv('download.folder/unzipped/fields.csv')
  names(plays) <- fields %>% pull(Header)
  
  return(plays)
}

plays2017.raw <- get_plays(2017)
```

As is, our dataframe contains dozens of columns we don't want and is missing a few that we do. First off, the analysis is much more intuitive if we have the text description of a play instead of the `EVENT_CD` encoding (i.e. `"Home run"` instead of `23`). More importantly though we need to know two things:

+ how many runs scored on the play
+ what the base-out situation (which bases are occupied, how many outs there are) is at the beginning and end of each play

To create the first column, we tally up the number of runners who advanced past third base (no this isn't quivalent to sayint `== 4` since in retrosheet's uses `5` and `6` in these columns to denote unearned runs).

To create the second column, we note whether each base is empty (`is.na`), and then `paste` these results together into a single `base` variable, which itself is combined with `out` to provide full situational information. 

We also need the base-out situation once the play is over, which we get from taking the base-out situation of the subsequent play with the `lead` function. Grouping by the game, inning number, and inning side ensures that the last play of the frame doesn't inherit empty bases, but rather the default that we define (`"END"`). 

```{r}
add_columns <- function(plays.raw){
  
  event_key <- read_csv('download.folder/unzipped/event_key.csv')
  # rewrite janky character
  event_key$RESULT[event_key$EVENT_CD == 19] <- "Fielder's Choice"
  plays.raw <- plays.raw %>% inner_join(event_key, by = "EVENT_CD") %>%
    mutate(RESULT = as.factor(RESULT))
  
  plays <- plays.raw %>% 
    rowwise() %>%
    # how many people scored on the play?
    mutate(runs.play = sum(BAT_DEST_ID > 3,
                           RUN1_DEST_ID > 3,
                           RUN2_DEST_ID > 3,
                           RUN3_DEST_ID > 3,
                           na.rm = TRUE)) %>%
    mutate(on.first = ifelse(is.na(BASE1_RUN_ID), '_', '1B'),
           on.second = ifelse(is.na(BASE2_RUN_ID), '_', '2B'),
           on.third = ifelse(is.na(BASE3_RUN_ID), '_', '3B')) %>%
    mutate(base = paste(on.first, on.second, on.third),
           out = OUTS_CT) %>%
    mutate(base_out = paste(base, out)) %>%
    group_by(GAME_ID, INN_CT, BAT_HOME_ID) %>%
    mutate(base_out_end = lead(base_out, default = "END"),
           HOME_TEAM_ID = substr(GAME_ID, 1, 3)) %>%
    select(GAME_ID, HOME_TEAM_ID, AWAY_TEAM_ID, INN_CT, BAT_HOME_ID, 
           OUTS_CT, AWAY_SCORE_CT, HOME_SCORE_CT, BAT_ID, PIT_ID, 
           EVENT_CD, AB_FL, H_FL, SH_FL, SF_FL, RESULT, 
           runs.play, base_out, base_out_end)
  
  return(plays)
}

plays2017 <- add_columns(plays2017.raw)
```

Adding the columns allows us to construct a **run expectancy matrix**, an important concept in sabermetrics. A run expectancy matrix gives us the average remaining runs to be scored in an inning based on the base-out situation. The highest value of about 2.2 runs occurs, as you'd expect, with bases loaded and nobody out, while the lowest value, about a tenth of a run, occurs with two outs and nobody on.

Run expectancy matrices are important for a variety of reasons. First of all they allow every play to be interpreted in the view of increased or decreased runs, which is the foundation of sabermetrics in the first place. As a result, you're forced to understand baseball as a procession of state transitions authored by the pitcher and batter, where the best players are those who can harvest more runs out of the states they inherit. Plus, they're also just a good reference for fans, since you can actually attach a number to your feelings about an inning as it unfolds: yes that leadoff double was good, but how good? Approximately two-thirds of a run.

Making such a matrix is mostly just an act of accounting. To calculate how many runs are left in an inning, you need to know two things: the current score and the score once the side is retired. To calculate the final score of an inning, we build a helper dataframe, `inning.runs` that keeps `runs.cum`, a cumulative sum of runs scored by each team within a game, and therefore acts as the scoreboard at the end of each half inning.

We then `inner_join` this dataframe to the original plays by the half inning, and calculate the `runs.left` as the difference between `runs.current` (readily available from the existing columns) and `runs.cum`. 

```{r}
make_RE_matrix <- function(plays){
  inning.runs <<- plays %>%
    group_by(GAME_ID, INN_CT, BAT_HOME_ID) %>%
    summarize(runs.inning = sum(runs.play)) %>% 
    arrange(GAME_ID, BAT_HOME_ID, INN_CT) %>%
    group_by(GAME_ID, BAT_HOME_ID) %>%
    mutate(runs.cum = cumsum(runs.inning))
  
  runs.league <<- inning.runs %>%
    group_by(GAME_ID, BAT_HOME_ID) %>%
    summarize(runs = max(runs.cum)) %>%
    ungroup() %>%
    summarize(runs = sum(runs)) %>%
    pull(runs)
  
  RE.matrix <- plays  %>%
    inner_join(inning.runs, by = c('GAME_ID', 'INN_CT', 'BAT_HOME_ID')) %>%
    mutate(runs.current = ifelse(BAT_HOME_ID == 1, HOME_SCORE_CT, AWAY_SCORE_CT)) %>%
    mutate(runs.left = runs.cum - runs.current) %>%
    group_by(base_out) %>% summarize(runs.expected = mean(runs.left)) %>%
    rbind(data.frame(base_out = "END", runs.expected = 0))
  
  return(list("matrix" = RE.matrix, "runs" = runs.league))
}

RE.matrix2017 <- make_RE_matrix(plays2017)
```

RUN VALUES BLAH

```{r}
# get_run_values <- function(plays, RE.matrix){
#   
#   matrix <- RE.matrix$matrix
#   runs <- RE.matrix$runs
#   
#   run.values <- plays %>%
#     inner_join(matrix, by = "base_out") %>%
#     inner_join(matrix, by = c("base_out_end" = "base_out")) %>%
#     rename(runs.expected = runs.expected.x,
#            runs.expected.end = runs.expected.y) %>%
#     group_by(RESULT) %>%
#     summarize(n = n(),
#               run.value = mean(runs.play + runs.expected.end - runs.expected)) 
#   
#   out.value <- run.values %>% 
#     filter(RESULT %in% c("Generic out", "Strikeout", "Fielder's Choice")) %>%
#     summarize(run.value = weighted.mean(run.value, w = n)) %>%
#     pull(run.value)
#   
#   run.values <- run.values %>%
#     mutate(run.value.over.out = run.value - out.value)
#   
#   PA.league <- sum(plays$AB_FL) + sum(plays$SH_FL) + sum(plays$SF_FL) +
#               sum(plays$RESULT %in% c("Walk", "Intentional walk", "Hit by pitch", "Interference"))
#   
#   runs.per.PA <- runs/PA.league
#   
#   wOBA.league <- (sum(plays$H_FL > 0) + sum(plays$RESULT %in% c("Walk", "Hit by pitch")))/
#                (sum(plays$AB_FL) + sum(plays$SF_FL) + sum(plays$RESULT %in% c("Walk", "Hit by pitch")))
#   
#   wOBA.raw <- run.values %>%
#     filter(RESULT %in% c("Home run", "Triple", "Double", "Single", "Hit by pitch", "Walk", "Intentional walk")) %>%
#     summarize(run.value.over.out = sum(run.value.over.out*n)) %>%
#     pull(run.value.over.out)/PA.league
#   
#   wOBA.scale <- wOBA.league/wOBA.raw
#   
#   run.values <- run.values %>% 
#     mutate(weight = run.value.over.out*wOBA.scale)
#   
#   return(list("RV" = run.values, "wOBA.league" = wOBA.league, "wOBA.scale" = wOBA.scale, "RpPA" = runs.per.PA))
# }
# 
# run.values.2017 <- get_run_values(plays2017, RE.matrix2017)
# ```
# 
# wOBA function rewrite
# 
# ```{r}
# get_wOBA <- function(player, plays, run.values){
#   
#   plays.player <- plays %>% ungroup() %>%
#     filter(BAT_ID == player)
#   
#   player.events <- plays.player %>% count(RESULT) %>% complete(RESULT, fill = list(n=0))
#   
#   NIBB <- player.events %>% filter(RESULT == "Walk") %>% pull(n)
#   IBB <- player.events %>% filter(RESULT == "Intentional walk") %>% pull(n)
#   HBP <- player.events %>% filter(RESULT == "Hit by pitch") %>% pull(n)
#   B1 <- player.events %>% filter(RESULT == "Single") %>% pull(n)
#   B2 <- player.events %>% filter(RESULT == "Double") %>% pull(n)
#   B3 <- player.events %>% filter(RESULT == "Triple") %>% pull(n)
#   HR <- player.events %>% filter(RESULT == "Home run") %>% pull(n)
#   SF <- sum(plays.player$SF_FL)
#   AB <- sum(plays.player$AB_FL)
#   PA <- sum(plays.player$AB_FL) + sum(plays.player$SH_FL) + sum(plays.player$SF_FL) +
#     sum(plays.player$RESULT %in% c("Walk", "Intentional walk", "Hit by pitch", "Interference"))
#   
#   RV <- run.values$RV
#   
#   wB1 <- RV %>% filter(RESULT == "Single") %>% pull(weight)
#   wB2 <- RV %>% filter(RESULT == "Double") %>% pull(weight)
#   wB3 <- RV %>% filter(RESULT == "Triple") %>% pull(weight)
#   wHR <- RV %>% filter(RESULT == "Home run") %>% pull(weight)
#   wBB <- RV %>% filter(RESULT == "Walk") %>% pull(weight)
#   wHBP <- RV %>% filter(RESULT == "Hit by pitch") %>% pull(weight)
#   
#   wOBA <- (wBB*(NIBB) + wHBP*HBP + wB1*B1 + wB2*B2 + wB3*B3 + wHR*HR)/
#           (AB + NIBB + SF + HBP)
#   
#   return(c('PA' = PA, 'wOBA' = round(wOBA, 3)))
# }
# 
# get_league_wOBAs <- function(plays, run.values,
#                              AL = c('BAL', 'BOS', 'CHA', 'CLE', 'DET', 
#                                     'HOU', 'KCA', 'ANA', 'MIN', 'NYA', 
#                                     'OAK', 'SEA', 'TBA', 'TEX', 'TOR')){
#   pitchers <- plays %>%
#     filter(INN_CT < 8) %>%
#     pull(PIT_ID) %>% unique()
#   
#   plays <- plays %>%
#     filter(!BAT_ID %in% pitchers) %>%
#     mutate(BAT_ID = ifelse(HOME_TEAM_ID %in% AL, "AL", "NL"))
#   
#   wOBA.AL <- get_wOBA("AL", plays, run.values)
#   wOBA.NL <- get_wOBA("NL", plays, run.values)
#   
#   return(
#     data.frame(
#       league = c("AL", "NL"),
#       wOBA = c(wOBA.AL['wOBA'], wOBA.NL['wOBA']),
#       PA = c(wOBA.AL['PA'], wOBA.NL['PA'])
#       )
#     )
# }
# 
# league.wOBAs.2017 <- get_league_wOBAs(plays2017, run.values.2017)
# 
# 
# make_park_factors <- function(plays, regress = 0.1){
#   park.factors = plays %>%
#     group_by(GAME_ID, HOME_TEAM_ID, AWAY_TEAM_ID) %>%
#     summarize(runs = sum(runs.play)) %>%
#     ungroup() %>%
#     select(-GAME_ID) %>%
#     rename(home = HOME_TEAM_ID, away = AWAY_TEAM_ID) %>%
#     gather(location, team, -runs) %>%
#     group_by(team, location) %>%
#     summarize(RPG = mean(runs)) %>%
#     spread(location, RPG) %>%
#     mutate(PF = (home/away)/2 + 0.5) %>%
#     mutate(PF = round((1 - regress)*PF + regress, 2)) %>%
#     select(team, PF)
#   
#   return(park.factors)
# }
# 
# park.factors.17.1yr <- make_park_factors(plays2017)
# 
# plays2016 <- add_columns(get_plays(2016))
# plays2015 <- add_columns(get_plays(2015))
# park.factors.17.3yr <- rbind(plays2015, plays2016, plays2017) %>%
#   make_park_factors(regress = 0.05)
# 
# get_wRC_plus <- function(player, plays, run.values, league.wOBAs, park.factors,
#                          AL = c('BAL', 'BOS', 'CHA', 'CLE', 'DET', 
#                                 'HOU', 'KCA', 'ANA', 'MIN', 'NYA', 
#                                 'OAK', 'SEA', 'TBA', 'TOR')){
#   
#   plays.player <- plays %>% ungroup() %>%
#     filter(BAT_ID == player)
#   
#   tm <- plays.player %>%
#     filter(BAT_HOME_ID == 1) %>%
#     slice(1) %>% pull(HOME_TEAM_ID)
#   
#   PF <- park.factors %>%
#     filter(team == tm) %>%
#     pull(PF)
#   
#   wOBA <- get_wOBA(player, plays, run.values)
#   PA <- wOBA['PA']
#   wOBA <- wOBA['wOBA']
#   
#   OBP <- run.values$wOBA.league
#   wOBA.scale <- run.values$wOBA.scale
#   RpPA <- run.values$RpPA
#   
#   lg <- ifelse(tm %in% AL, 'AL', 'NL')
#   wOBA.league <- league.wOBAs %>% filter(league == lg) %>% pull(wOBA)
#   PA.league <- league.wOBAs %>% filter(league == lg) %>% pull(PA)
#   
#   wRC.league <- ((wOBA.league - OBP)/wOBA.scale + RpPA)*PA.league
#   
#   wRC.plus <- 100 * ( ( (wOBA - OBP)/wOBA.scale + RpPA) + (RpPA - PF*RpPA) )  /
#                     (wRC.league/PA.league)
#   
#   return(round(unname(wRC.plus)))
# }
# 
# 
# # get_wRC_plus('blacc001', plays2017, run.values.2017, league.wOBAs.2017, park.factors.3yr)
# # get_wOBA('judga001', plays2017, run.values.2017)
# # get_wRC_plus('judga001', plays2017, run.values.2017, league.wOBAs.2017, park.factors.2017)
# # get_wRC_plus('stanm004', plays2017, run.values.2017, league.wOBAs.2017, park.factors.2017)
# 
# players.2017 <- read_csv('/Users/walkerharrison/Downloads/master.csv') %>%
#   mutate(playerid = as.integer(fg_id))
# 
# fg <- read_csv('/Users/walkerharrison/Downloads/FanGraphsLeaderboard.csv')
# 
# leaders <- fg %>% inner_join(players.2017, by = 'playerid') %>%
#   select(playerid, retro_id, Name, wOBA, `wRC+`)
# 
# leaders <- leaders %>% rowwise() %>% 
#   mutate(wOBA.us = get_wOBA(retro_id, plays2017, run.values.2017)['wOBA'],
#         wRC.plus = get_wRC_plus(retro_id, plays2017, run.values.2017, league.wOBAs.2017, park.factors.17.3yr))
# 
# g1 <- leaders %>% ggplot(aes(wOBA, wOBA.us)) + geom_point(size = 2, alpha = 0.3) + 
#   geom_abline(slope = 1, intercept = 0) + theme_bw()
# 
# g2 <- leaders %>% ggplot(aes(`wRC+`, wRC.plus)) + geom_point(size = 2, alpha = 0.3) + 
#   geom_abline(slope = 1, intercept = 0) + theme_bw()
# 
# grid.arrange(g1, g2, nrow = 1)
```